---
title: "Measuring Meaning in Mixed Methods - Week 3"
output:
  github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Topic Modeling using Latent Semantic Analysis

```{r}
#We construct the example matrix
human <- c(1,0,0,1,0,0,0,0,0)
interface <- c(1,0,1,0,0,0,0,0,0)
computer <- c(1,1,0,0,0,0,0,0,0)
user <- c(0,1,1,0,1,0,0,0,0)
system <- c(0,1,1,2,0,0,0,0,0)
response <- c(0,1,0,0,1,0,0,0,0)
time <- c(0,1,0,0,1,0,0,0,0)
EPS <- c(0,0,1,1,0,0,0,0,0)
survey <- c(0,1,0,0,0,0,0,0,1)
trees <- c(0,0,0,0,0,1,1,1,0)
graph <- c(0,0,0,0,0,0,1,1,1)
minor <- c(0,0,0,0,0,0,0,1,1)
X <- rbind(human,interface,computer,user,system,response,time,EPS,survey,trees,graph,minor)
colnames(X) <- c("c1","c2","c3","c4","c5","m1","m2","m3","m4")
print(X)
```

Measure the intial correlation between several rows

```{r}
cor(human,user, method="spearman")
cor(human,minor, method="spearman")
```

Do the singular value decomposition on the X matrix

```{r}
SVD <- svd(X)
print(SVD)
rownames(SVD$u) = rownames(X) 
rownames(SVD$v) = colnames(X)
u <- as.matrix(SVD$u)
v <- as.matrix(SVD$v)
d <- SVD$d
```

We can set all but the first two singular values to zero. this means that r=2

```{r}
d[3:9] <- 0
d_r <- as.matrix(diag(d))
print(d_r)
```

We put the matrix back together but now with only keeping the first two singular values.

```{r}
X_r <- u %*% d_r %*% t(v)
print(round(X_r,2))
```

We measure the correlations among the rows(=the words)

```{r}
cor(t(X_r), method="pearson")
cor(t(X_r), method="spearman")
```

We measure the correlations among the columns(=the documents)

```{r}
cor(X,method="pearson")
cor((X_r), method="pearson")
```

The differences between the different topics should now be more clearly defined.





